### 2018-02-02 thesis/contents/discussions.tex
Updates from ShareLaTeX

\chapter{General Discussions}
\section{Tractography}
\subsection{Delineating the trigeminal nerve into the brainstem}
The primary motivation to visualize the trigeminal nerve via tractography, was due to the inability to discern both the brainstem trigeminal nerve, nor the trigeminal nucleus in all known MRI modalities. Before tractography, the curvature of the trigeminal nerve could only be inferred from previous anatomical studies, or by postmortem dissection. 
For successful tractography of the CN V, both DWI sequence and the algorithm choice is paramount. Given a DWI sequence, at 25 gradient directions, cisternal CN V could be readily reconstructed with Gaussian tensors. Resolution of the DWI sequence would affect the output. Unlike major white matter bundles, cranial nerves diameters are in the milimeter range, and therefore voxel sizes greater than 2 mm will result in severe partial volume of cranial nerves. Achieving 2mm isovoxel on a 3T scanner requires 32 channel head coil. Standard clinical 3T MRI scanners with 8 channels head coils can only achieve 2.6 mm isovoxel resolution before acquisition time and noise becomes insurmountable. For this reason, the acquisition in-plane resolution can be tune to 1x1 mm, and slice thickness at 3mm. While the acquisition is not ideal, it sacrifices resolution in only the z-axis, the tractography visual fidelity is improved. 
Discerning brainstem CN V requires algorithms that can reliably resolve tract structures in regions with crossing fibers. A number of methods have been devised, including probabilistic tractography made popular by the FSL software package, which generates image-based results that represent the propagation probability of a seed region. Neurosurgeons however often need visualizations that represent anatomy geometrically. Therefore streamline-based tractography visualizations are of broader appeal. In our experience, both XST and sphere harmonic based propagation algorithms can perform adequately to delineate brainstem CN V, as long as there are greater then 50 gradient directions. 
Another often overlooked area is the seeding pattern. When tracing the streamlines, an algorithm will need to generate a number of starting points in a defined volume. Common patterns in prepackaged software include in a regularly spaced grid, where seed spacing can make decisive differences in visuals, and random sampling. The seeding pattern is important in both the final visual style of the tracts, as well as if key anatomical landmarks can be properly tracked. Another consideration is the reproducibility of the desired anatomy, as randomly seeded bundles cannot be reproduced exactly. Grid seeding also can produce sparse spacing of the tract bundles when viewed from a certain angle, while random seeding can reduce this kind of visual artifact. A good balance might be to have a reproducible sampling scheme that also is robust to different viewing directions, for example a phyllotaxis pattern.

\subsection{Tractography group registration}
Tractography is a powerful methodology that reveals the morphology of brain white matters that were otherwise previously invisible. Its primary usage however has been limited to case studies and small-scale anatomy comparative analysis. The continuous advancements in big data and machine learning demand the adaptation of new neuroimaging methods to the analysis and quantification of large groups.

Group comparative studies require the establishment of anatomical consensus, such that similar white matter structures between populations can be readily studied. Efforts in MRI medical image registration have paved the way to this end, especially in the fields of diffeomorphic registration. The difficulty in adapting registration to tractography is that tractography data are 3D geometries, and therefore are extremely difficult for analytical methods. The best efforts to date apply affine transformation for between-tract registrations \cite{Garyfallidis2015}. However non-linear geometric registration is still a topic of future research. Our approach is to recast tractography as a brain segmentation method that is spatially-aware. Tractography then can be considered a point-based sub-sampling method that is represented in the same coordinate frame as the native DWI space. Any transforms applied to this DWI space is then applicable also to any tractography models that share its coordinate frame. Existing diffeomorphic registration solutions can then be used to non-linearly distort tractography, and circumvent the computational difficulty of geometry based non-linear registration. The advantage of our tractography registration strategy is that any new advances in the field of MR registration can be readily adapted to improve the output of our method.

Group registration of tractography towards a common template space requires two separate steps: T1---DWI co-registration, and T1---Template registration. Any registration errors in these two steps will become multiplicative in the end result. T1 registration is well studied, however T1---DWI registration has few literature. Therefore it was imperative that we investigated the best method to minimize registration errors that were compatible with our image data format.

\subsection{Measurements from tracts}
Aggregating along-the-measure statistics is a form of clustering that requires projection of tract vertices onto a structural skeleton, or centroid. Common clustering methods calculate a polyline centroid based on different distancing functions between streamlines. These methods,  however, require tractography streamlines to have similar length, and will underperform when tractography structures contain sharp turns. We have explored a number of strategies, and considered adaptation of manifold-based methods, and point-cloud skeletal extraction method from computer graphics literature. These approaches, however, contain assumptions that do not suit our needs. For example, manifolds commonly consider that the latent structure of the data lies on a lower-dimensional surface, while tractography is commonly arranged in complex bundles. Moreover, point-cloud methods deal with point-cloud data obtained from 3D scanners, which situate on a 3D shell, while tractography points are distributed within a volume. White matter anatomy can have intricate crossings, multi-way join and diverging structures. These are difficult for existing 3D analytical methods, and require models with simplified assumptions to reduce the number of unknown parameters. A potentially promising future direction may be to automatically discover these more complex latent structures using deep learning methods.

@@

The findings in Study I \ref{section:study1} revealed localized CN V tissue differences in MS-TN and idiopathic TN. The possible implication is that disruption anywhere along the primary trigeminal afferent may lead to similar TN-like pain syndromes. MS is a classic myelin inflammatory disease, and therefore these diseases may share similar disruptions in myelin ultrastructure.

Study IV further suggested that the unaffected CN V diffusivity is also different from that of controls. The unaffected side was previously presumed to be similar to normal tissue due to the absence of pain. In many studies it has also been used as a basis of comparison. This finding however is not as surprising as one may suspect. It has been long observed that bilateral neurovascular compression could be observed in some unilateral TN cases. Moreover, in third-degree neurovascular compression, severe vascular hardening and deformation can be observed in the cisternal space. It is reasonable that these observable macroscopic changes can effect CNS tissue changes on the unaffected side. Moreover descending modulation from PAG can become maladaptive or ineffective in chronic pain, and in these cases more centrally involved changes in the trigeminal nucleus and its associated pathways are possible.

@@

>>>>

\chapter{General Discussions}

\section{Summary Of Findings}

The contribution of this thesis are as follows:

\begin{itemize}

\item We are the first to image the brainstem segment of the CN V \textit{in vivo} using multi-tensor diffusion tractography. Prior to this work, CN V tractography was limited to single-tensor DTI models and cannot be delineated into the pons. Clear delineation of the pontine CN V nerve allows reliable measurements of CN V diffusivity, therefore greatly increase the range of trigeminal-related pathology, for example MS-TN, that can be studied with neuroimaging.

\item The study, for the first time, clearly differentiated TN and MS-TN by sub-segment diffusivity differences, between cistern (TN) and brainstem peri-lesional (MS-TN) segments. This distinction in the trigeminal REZ between these two conditions have been suspected, but never confirmed. The study is the first to suggest that disruptions along the CN V in different locations may trigger similar pain symptoms.

\item We overcame the major technical barrier to detailed diffusion tractography study of the trigeminal pathway in a large population group, by developing and validate the Selective Automated Group Integrated Tractography (SAGIT) software platform. In the process, we determined the optimal T1-DWI co-registration methodology with clinical DWI sequences. Prior to this project, the co-registration method has never been exhaustively determined and reported in literature.

\item We are the first to perform merged group tractography study of the trigeminal nerve, and its primary sensory pathway, and the first to conduct such study in TN, and the first to perform Gaussian Process machine learning to the study of TN. We studied the trigeminal primary (CN V), secondary (trigeminopontothalamic, TPT), and tertiary (trigeminocortical, S1) white patter pathways in its cross-sectional detail. We discovered that the unaffected CN V diffusivity differs from controls as well as the affected side and that the affected TPT pathway and S1 pathway diffusivity differs from controls.

\end{itemize}

@@

The findings in Study I (Chapter \ref{section:study1}), which aimed to identify trigeminal nerve diffusivity signatures in TN and MS-TN, identified for the first time a method that can distinguish between microstructural changes in these conditions, and revealed localized CN V diffusivity differences in MS-TN and idiopathic TN. The possible implication is that disruption anywhere along both the cisternal and pontine segments of the trigeminal nerve may lead to similar TN-like pain syndromes. Central demyelination in the trigeminal root transition zone is associated with vascular compression in TN. Moreover MS is a classic myelin inflammatory disease. Therefore both diseases may share pathophysiology due to disruptions in myelin ultrastructure. The study opened the possibility that multi-tensor tractograpy aided analysis added a measure of objective assessment in examining pain conditions, and may facilitate similar studies that provide insight into the fundamental changes that occur at the level of white matter.

Study IV (Chapter \ref{section:study4}), which applied group tractography to the study of the trigeminal sensory pathway, further suggested that the unaffected CN V diffusivity is also different from that of controls. DeSouza et al. \cite{Desouza2013} also presented similar findings, where the unaffected RD, MD, and AD were found to differ from controls. Desouza performed the study with manual ROI placement at a single measurement position in the REZ, Whereas we performed GP classification along the nerve automatically. The unaffected side was previously presumed to be similar to normal tissue due to the absence of pain. In many studies it has also been used as a basis of comparison. This finding, however, is not surprising. It has been long observed that bilateral neurovascular compression could be observed in some unilateral TN cases. Moreover, in third-degree neurovascular compression, severe vascular hardening and deformation can be observed in the cisternal space. It is reasonable that these observable macroscopic changes can effect CNS tissue changes on the unaffected side. Moreover descending modulation from PAG can become maladaptive or ineffective in chronic pain, and in these cases more centrally involved changes in the trigeminal nucleus and its associated pathways are possible. With this additional insight, it becomes important to not interpret results of affected-to-unaffected comparisons in terms of differences between affected and healthy controls. Additionally, the logical next question is to determine that, if the unaffected nerve differs from controls, at what degree of difference does it result in pain. Investigations into bilateral TN pain, and vascular compression in the absence of pain, may reveal additional insight into the relationship between white matter changes and TN pain in general.

@@

\section{Tractography of the trigeminal nerve and brainstem fibers}

The primary motivation for the study of the trigeminal nerve with tractography was the significant limitation of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus. Prior to the use of tractography, the structure of the trigeminal nerve could only be inferred from direct anatomical studies or postmortem dissection.

Resolving cross fibers of the cerebral white matter was a fundamental advancement in diffusion tractography. The brain white matter was believed to consists of 90\% complex crossing pathways, and the inability for DTI to reliably image these structure posed a fundamental challenge for neuroimaging. 
A widely adopted technique was probabilistic tractography, which used tissue priors to sample a large number of directions at each step stochastically, and combined the final result as a volumetric probability map. This kind of image-based probabilistic tractography was useful for whole-brain white matter studies. However, it was constrained to the resolution of the diffusion acquisition, which was insufficient for localized and small diameter pathways. Moreover, the algorithm depended on pre-generated tissue-priors, which excluded cranial nerves. 
Deterministic algorithms such as diffusion spectral imaging (DSI), and the related multi-shell acquisition strategies attempted to model the fiber orientation distribution by dramatically increasing the number of gradient directions (256--512 directions were required) in the diffusion acquisition. The acquisition time of these sequences was impractical for the clinical environment. Reduction in acquisition time for these algorithms is currently an active area of research \cite{}.
The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), were a significant development. SH-based methods modelled the fiber orientation distribution as a decomposition of SH basis functions. This meant that the methods are more robust to noise, and therefore require less number of gradient acquisitions. CSD proved to be particularly robust to noise, and can be deployed on acquisitions with only 55 directions. Another family of tractography (XST) algorithm constrained its model assumption to two fiber crossings, also proved to provide good results with 50 gradient directions.

These key algorithmic developments decidedly brought cross-fiber tractography in clinical neuroimaging to the realm of possibility. This is significant as the vast majority of imaging of neuropathologies in patients only exist as clinical acquisitions. We took advantage of this opportunity and explored the feasibility of imaging the pontine trigeminal nerve with CSD and XST. The trigeminal nerve is an ideal imaging target, both due to our immediate interest in TN, and also that the trigeminal nerve has a number of unique properties as a tractography target: 1) delineation quality is easy to judge due to its simple morphology. 2) it situates in both single-fiber (cistern) and crossing-fiber environments; 3) it is the largest cranial nerve, and therefore a good benchmark for imaging small anatomy.

The ability to resolve the pontine trigeminal nerve into the trigeminal nucleus was encouraging. It signified that for the first time we could segment the pontine trigeminal anatomy reliably with computer aid. Previously its identification was done by eye, and sometimes with simple heuristics based on distance from surrounding landmarks. Such manual methods were highly subjective and became problematic with patient images, where distortions to the nerve could not be assessed.

The ability to segment the entire trigeminal nerve also opens the avenue to measure its tissue diffusivity from tractography directly. Moreover, specific tissue-metrics such as myelin image, or PET images, can also be measured with confidence in reproducibility. This extends these methods to facilitate disease studies well beyond TN. We were able to divide the trigeminal nerve into 30 discrete cross-sections along its path, and observed as never before, detailed diffusivity changes as the nerve transits from pontine tissue to the cistern. Determining a common centroid curve from many subjects was computationally non-trivial, especially when the streamline lengths were highly variable, due to the noise in the cistern CSF. Turns out that the merged tractography strategy simplified this task by allowed a single manual curve to be defined manually in the template tractography space. This strategy can similarly be extended for obtaining along-the-track measure for other highly variable streamline bundles.

We applied machine learning to the analysis of TN in Study IV in order to avoid our own confirmation bias. The analysis of the trigeminal nerve in TN versus controls served as validation for the study's method. We had performed traditional multiple comparison statistical analysis, but found the result hard to interpret without inducing biases. The GP classifier was chosen, over more popular SVM classifiers, for the following reasons: 1) the obtained metrics show high location dependency. 2) SVM results are hard to interpret, whereas GP is a bayesian method, and is much more intuitive.

\section{Neuroimaging of deeper white matter for the study of pain}

The pain neuromatrix, as studied by various functional and structural connectivity studies, distributes over the full range of neocortical areas. These areas are associated with complex white-matter pathways that extends to all regions of the brain. These include sharp curvatures of the anterior pain pathways of the ACC, the distant projections of the temporal cortices. Some pathways are just beyond the feasibility of current methods, these include the projections of the insula and claustrum, small decending pathways to PAG, and the spinal trigeminal pathways in the brainstem.

We have shown in Study III (\ref{section:study3}), for the first time, that different tractography algorithms are suited to different neuropathways when comparing merged reconstructions in 42 subjects. Maier-Hein et al came to similar conclusions after examining 96 reconstructions from a single set of simulated data \cite{Maier-Hein2017}.
Therefore care must be taken to ensure tractography parameters are targeted to the anatomy of study, as more consistent delineation lead to more robust tissue measurements.

We observed that merged tractography can delineate more complete pathway delineations than from any single subject. This is similar to the observation that averaged MRI anatomical images from multiple subjects can produce high quality templates. Therefore merged tractography may serve to produce high quality tractography templates, that can form the basis for both geometric and measurement comparisons in individuals. We have explored the use of SAGIT in generating the first neonatal porcine tractography template \cite{Zhong2016a}. Such template does not exist for humans to the best of our knowledge, and can be readily created from Human Connectome Project data. In patients where tractography cannot be reliably delineated, for example in the case of severe white matter degeneration in MS or low resolution clinical acquisitions, these template tracts may be used as proxy to enable analysis.

\section{Neurosurgery and Tractography}

The biggest value of tractography to neurosugery is to assit in pre-operative planning. Tractography, in conjuction with 3D anatomical segmentation, can 1) reconstruct white matter pathways that are otherwise invisible, and 2) allow intuitive spatial relationship of key anatomies that are hard to visualise in 2D. We had previously demonstrated these capabilities in the reconstruction of deformed facial/vestibular nerve complex in relation to vestibular schwannomas \cite{Chen2011b,Behan2017}. Tractography and 3D anatomical model ultimately are standard 3D geometries in computer graphics. Therefore advances in 3D printing can be easily adopted in neurosurgery to allow surgeons to physically examine tractography reconstructions. On the otherhand, technologies such as virtual reality (VR) and augmented reality (AR), which are simply different ways to present computer graphics, can allow new ways to empower surgeons in decision making. Tractography can play a key role in bridging computer vision based diagnosis and surgical disease intervention to improve surgical outcomes and improved quality-of-life for patients.

\section{Method Discussions}

\subsection{Delineating the brainstem fibers of the trigeminal nerve}

For successful tractography of the CN V, both DWI sequence and the algorithm choice are paramount. Given a DWI sequence, cisternal CN V could be readily reconstructed with Gaussian tensors. Unlike major white matter bundles, cranial nerves diameters are in the millimetre range, and therefore voxel sizes greater than 2 mm may result in partial volume averaging. Achieving 2mm isovoxel on a 3T scanner requires 32 channel head coil. Standard clinical 3T MRI scanners with 8 channels head coils can only achieve 2.6 mm isovoxel resolution before acquisition time and noise becomes insurmountable. For this reason, the acquisition in-plane resolution can be tuned to 1x1 mm, and slice thickness at 3mm. While the acquisition is not ideal, it sacrifices resolution only in the z-axis, however the tractography visual fidelity is improved. 
Imaging the brainstem CN V requires algorithms that can reliably resolve tract structures in regions with crossing fibers. A number of methods have been devised, including probabilistic tractography (FSL)\cite{Behrens2007}, which generates image-based results that represent the propagation probability of a seed region. Neurosurgical visualization however requires three dimensional representation of the fibers, and therefore streamline-based tractography visualizations are of broader appeal. In our experience, both XST and sphere harmonic based propagation algorithms can perform adequately to delineate brainstem CN V, as long as there are greater then 50 gradient directions.

An important technical point, but rarely mentioned in literally, is the tractography seeding pattern. Therefore I will address it here: When tracing the streamlines, an algorithm will need to generate a number of starting points in a defined volume. Common patterns in prepackaged software include in a regularly spaced grid and random sampling. The seeding pattern is important in both the final visual representation of the tracts, and to determine whether key anatomical landmarks can be properly tracked. Another consideration is the reproducibility of the desired anatomy, as randomly seeded bundles cannot be reproduced deterministically. Grid seeding may result in sparse spacing of the tract bundles when viewed from a certain angle, while random seeding can reduce this kind of visual artefact. A good balance might be to have a reproducible sampling scheme that also is robust to different viewing directions, for example a phyllotaxis pattern.

\subsection{Tractography group registration}
Tractography is a powerful methodology that permits the visualization of white matter fibers in three dimensions. Its primary use however has been limited to case studies and small-scale anatomy comparative analysis. The continuous advancements in big data and machine learning demand the adaptation of new neuroimaging methods to the analysis and quantification of larger groups.

Group comparative studies require the establishment of anatomical consensus, such that similar white matter structures between populations can be readily studied. Efforts in MRI medical image registration have paved the way to this end, especially in the fields of diffeomorphic registration. The difficulty in adapting registration to tractography is that tractography data are 3D geometries, and therefore are extremely difficult for analytical methods, as they are computationally untenable. To date there is no direct non-linear 3D deformation methods for tractography. The best efforts to date apply affine transformation for between-tract registrations \cite{Garyfallidis2015}. Our approach is to recast tractography as a brain segmentation method while also make use of its orientation information. Tractography then can be considered a point-based sub-sampling method that is represented in the same coordinate frame as the native DWI space. Any transforms applied to this DWI space is then applicable also to any tractography models that share its coordinate frame. Existing diffeomorphic registration solutions can then be used to non-linearly distort tractography, and circumvent the computational difficulty of geometry based non-linear registration. The advantage of our tractography registration strategy is that any new advances in the field of MR registration can be readily adapted to improve the output of our method.

Group registration of tractography towards a common template space requires two separate steps: T1---DWI co-registration, and T1---Template registration. Any registration errors in these two steps will become multiplicative in the end result. T1 registration is well studied, however T1---DWI registration has few literature. This demonstrates the need to investigate the best method to minimize registration errors that were compatible with our image data format.

\subsection{Measurements from tracts}
Aggregating along-the-measure statistics is a form of clustering that requires projection of tract vertices onto a structural skeleton, or centroid. Common clustering methods calculate a polyline centroid based on different distancing functions between streamlines. These methods require tractography streamlines to have similar length, and will underperform when tractography structures contain sharp turns. We have explored a number of strategies, and considered adaptation of manifold-based methods, and point-cloud skeletal extraction method from computer graphics literature. These approaches, however, contain assumptions that do not suit our needs. For example, manifolds commonly consider that the latent structure of the data lies on a lower-dimensional surface, while tractography is commonly arranged in complex bundles. Moreover, point-cloud methods deal with point-cloud data obtained from 3D scanners, which situate on a 3D shell, while tractography points are distributed within a volume. White matter anatomy can have intricate crossings, multi-way join and diverging structures. These are difficult for existing 3D analytical methods, and require models with simplified assumptions to reduce the number of unknown parameters. A potentially promising future direction may be to automatically discover these more complex latent structures using deep learning methods.

### 2018-02-02 thesis/contents/hypotheses.tex
Updates from ShareLaTeX

The general aim of this thesis is quantify the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways in order to identify specific diffusivity pattern that differentiates between TN and non-TN CN V tissue. In achieve this aim, a generalizable group diffusion imaging tractography software framework must be developed to overcome the technical barrier of large scale group tractography in arbitrary populations. A number of confounds on data registration and measure must be resolved in order to gain confidence in the result of this methodology. These include a) establish the feasibility of diffusion tractography of brainstem CN V originating from the trigeminal nucleus; b) establish the best way to minimize DWI to T1 co-registration error in order to allow multi-modal template space registration across multiple subjects; c) Make possible non-linear registration of tractography geometries, and to quantify the geometry registration performance at the group level; d) apply previously established methodology to the reconstruction and quantification of the trigeminal sensory pathway.

>>>>

The general aim of this thesis is to introduce quantitative measures for the study of trigeminal nerve microstructure in health and disease. This necessitates detailed \textit{in-vivo} microstructural analysis of the nerve and its associated pathways, to determine how the microstructure is altered in idiopathic TN, symptomatic TN as a result of MS, and healthy controls. With an increasing number of subjects, single subject analysis becomes cumbersome. Critical methodological advances are necessary for the feasibility of the general aim. These include the development a group diffusion tractography tool, determination of the most robust intra-subject imaging co-registration strategy, and validation of the group tractography output across a variety of tractography algorithms and white matter anatomy. Lastly, these methodological advances will be applied to the large-scale study of TN versus healthy control.  The more considerable amount of data output from the methodology will enable the application of machine learning towards the aim of quantifying the specific group-wise diffusivity changes along the trigeminal sensory white matter pathways, to identify specific diffusivity pattern that differentiates between TN and non-TN. The roadmap and rationale for the specific studies are described below.

\section{Study I: Diffusivity signatures characterize trigeminal neuralgia associated with multiple sclerosis}
MS-TN is thought to relate pathophysiologically to MS lesions in the brainstem, in the area of the trigeminal fibers in the pons. However the exact anatomical relationship between the lesions and trigeminal nerve within the brainstem has been difficult to quantify, due to the limited contrast and resolution in standard MRI and the variability of MS lesion across individuals. Previous single-tensor tractography studies of CN V have been limited to the imaging of the cisternal nerve segment. In this study, we use multi-tensor tractography with two aims: first, we focus on the delineation of the brainstem sub-segments of CN V; and the second, on the feasibility of applying multi-tensor tractography in the study of diffusivity differences in TN and MS-TN.

\subsection{Main Aim} 
Establish the feasibility of multi-tensor diffusion tractography in delineating the cisternal, root entry zone and brainstem segments of CN V and how these changes correlate with pain in MS-TN.

\subsection{Specific Aim}
\begin{itemize}
    \item Use multi-tensor tractography to delineate CN V into the the pontine brainstem, in order to quantify tissue diffusivity measurements in targeted sub-regions.
    \item Differentiate TN and MS-TN based on the obtained CN V diffusivity measurements.
\end{itemize}

\subsection{Specific Hypothesis}
\begin{itemize}
    \item TN is known to relate to vascular compression of the cistern REZ, therefore we expect its diffusivity signature to occur in the cisternal region.
    \item MS-TN should be related to brainstem MS lesions. Therefore CN V segments close to the lesions are expected to show more significant changes in tissue diffusivity, whereas segments far away from the lesions will not. 
    \item As a demyelinating disease, we expect to find a pattern of change in diffusivity signatures that reflects demyelination, including increase in RD, and decrease in FA. 
\end{itemize}

\section{Study II: Diffusion Weighted Image to T1 Co-registration Without Reversed-blip: Investigation of Best Practices}
The task of establishing inter-subject anatomical correspondences is imperative for extending tractography to multiple subjects. Population studies with inter-subject T1 registration are well-established, where a normalized T1 template is generated from numerous subjects to facilitate data analysis. Similarly, DWI of each subject can be co-registered to their corresponding T1, to extend the T1 template space to tractography space. Therefore the DWI-T1 co-registration accuracy is crucial to the reliability of inter-subject tractography alignment. DWI-T1 co-registration is challenging due to the DWI eddy-current and EPI field distortions. While new DWI sequences with reversed-blips have been shown to reduce non-linear distortions, they have yet to be commonly adopted in clinical practice. While T1 inter-subject registration methodology is well documented, the feasibility, and strategy for DWI-T1 co-registration on sequence without reversed-blips have not been documented, and forms a potential bottleneck on the performance of any DWI-T1 population registration pipeline.

\subsection{Main Aim}
Establish the best strategy to minimize DWI to T1 co-registration error in order to allow multi-modal template space registration across multiple subjects.

\subsection{Specific Aim}
\begin{itemize}
    \item Quantify the degree of improvement in registration between affine-only versus symmetric-diffeomorphic registrations.
    \item Determine which commonly-used DWI scalar image types can provide the best registration result when used as the registration-intermediate.
    \item Determine if anisotropic-power image, an experimental DWI scalar map that closely resemble T1 contrast, can provide even more registration improvements. 
\end{itemize}

\subsection{Specific Hypothesis}
\begin{itemize}
    \item Symmetric-diffeomorphic registration will result in the best accuracy, but with limited asymptotic improvement over affine registration alone.
    \item Averaged DWI should be the best scalar image, as it provides a balanced contrast between white and grey matter.
    \item Anisotropic-power (AP) image will show more substantial improvement over the existing scalar images.
\end{itemize}

\section{Study III: Merged Group Tractography Evaluation with Selective Automated Group Integrated Tractography}
The construction of an automated group tractography software framework depends on the establishment of the feasiblity of CN V brainstem delineation using multi-tensor tractography, and the optimal DWI-T1 co-registration strategy. The software aims to automate the tractography process on a large-scale, and also aims to streamline research workflow by providing region-of-interest, and tractography parameter management. It also provides the ability to report the tractography output at a group level both visually, as well as quantifying the output by an original scalar score called Normalized Overlap Score (NOS) score.

\subsection{Main Aim}
The aim of this study is to validate and demonstrate the software's flexibility by comparing the performance of four tractography algorithms across six anatomical regions in a healthy population. The tasks are performed  at a scale that would be prohibitive with manual methods. The comparisons determine the best tractography methods to use for different types of neuroanatomy.

\subsection{Specific Aim}
\begin{itemize}
    \item Compare DTI, XST, deterministic constrained-spherical deconvolution (CSD), and probabilistic CSD in 42 subjects. 
    \item Use the above techniques to study a series of white matter tracts that have been challenging to delineate in full detail, including: fornix, facial/vestibular-cochlear cranial nerve complex, vagus nerve, rubral-cerebellar decussation, optic radiation, and auditory radiation.
    \item Validate NOS scores by comparing merged tractography derived NOS scores with manual rating by neuroanatomy experts.
\end{itemize}

\subsection{Specific Hypothesis}
\begin{itemize}
    \item DTI is known to not able to delineate cross-fibers regions. If the group tractography works consistently, then its output should reflect such known algorithmic limitations, and DTI group tracts should not be able to perform well on more complex anatomies. The grouped tracts should however show consistent regional anatomical overlap across individuals.
    \item By merging the tracts of many individuals, the tractography delinations can complement each other, and this may improve the anatomical fidelity over individual cases.
    \item Specific tractography algorithms may be preferentially suited for the visualization of specific areas in the brain. For examples, deterministic tractography should preferentially be more suitable for nerve tractography, and probabilistic method better for deep neural pathways such as the optic and auditory radiations.
    \item NOS score should correlate with expert ratings, and show improved rating consistency over human rater variability.
\end{itemize}

\section{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
The developed methodology is applied to the study of the trigeminal sensory pathway. The method used is now thoroughly tested, and the best-suited tractography methods determined from the previous study. The study aims to examine the diffusivity signature differences in the trigeminal sensory pathway between TN and controls in a larger group. The pathway is divided into three major regions: CN V, trigeminopontothalamic (TPT) segment, and the thalamocortical S1 segment. With the increased datapoints, more detailed along-the-tract analysis can be carried out on each segment. Following this, we use machine learning algorithms to auto-determine relevant segments that maximally differentiates TN from controls.

\subsection{Main Aim}
The aim of this project is the reconstruction of the trigeminal pathways from the cisternal segment to the level of S1 sensory cortex, and examination using Gaussian Process classification.

\subsection{Specific Aim}
\begin{itemize}
    \item Delineate and visualize the three segments of the trigeminal sensory pathway: CN V, trigeminopontothalamic and  thalamocortical S1 segment. 
    \item At each segment, we will look for microstructural features to distinguish the affected vs. unaffected sides.
\end{itemize}

\subsection{Specific Hypothesis}
\begin{itemize}
    \item To validate the study, CN V diffusivity signature should match findings from Study I.
    \item At the level of CN V, the affected cistern segment should show differentiation from the unaffected side.
    \item The affected S1 segment should differ from the unaffected side.
    \item The affected TPT segment might show more differentiation near the end points, due to possible differences in trigeminal nucleus and thalamus diffusivity. However, its mid-segment diffusivity might be inconclusive, as the diffusivity metrics such as FA and RD might not be adequate in describing the complex crossings at the pontine decussation. 
\end{itemize}

### 2018-02-02 thesis/contents/reviews/mri.tex
Updates from ShareLaTeX

\subsection{Basic Principles}
\subsubsection{Spin and Magnetic Moment}
Spin is an intrinsic property of elementary particles, and can be either positive or negative. The spin quantum number  $s$ is defined as $ s = \frac{n}{2}$, where $n \in \{0,1,2,...\}$. Therefore $s$ is a quantized value in multiples of $\frac{1}{2}$. Protons, electrons and neutrons possess spin. Nuclei with odd number of protons or neutrons have non-zero nuclear spins; while those with even number of protons and neutrons have no spin. Particles with opposite spins can pair up to produce an overall spin of zero. Hydrogen nuclei, which makes up 63$\%$ the human body, possess spins of $\frac{1}{2}$. The spin of a particle has a magnetic moment vector that behaves like a magnet with north and south poles (Figure \ref{fig:spin}).

@@

There are more protons with spins with parallel orientation than anti-parallel, due to the increased energy stability. This number of difference is called spin excess and the magnetic moments of the excess spins add to form the net magnetization vector $M_z$. When the spin system is at equilibrium, net magnetization lies along the z axis, parallel to $B_0$.

@@

\subsubsection{Relaxation}
A transverse magnetization is introduced into the system that oscillates at radio frequency (RF) in order to modify the net magnetization, this is called the RF pulse. When exposing the nuclear spin system to the energy of the RF pulse, the equilibrium of the spin system is disturbed. The RF pulse is designed to introduce energy E such that the net spin direction orients at $90^\circ$. The net magnetization now is denoted as $M_{xy}$ as it aligns with the xy-plane instead of the z-axis. Individual spins now precess in phase with the RF pulse, it gives rises to the MR signal in the receiver coil.

With the termination of the RF pulse, the nuclei in the excited state will return to the equilibrium condition, this process is called \textit{relaxation}. 
The energy transmitted decays rapidly by two relaxation processes:  spin-lattice and spin-spin relaxations. These are responsible for \textit{T1 relaxation} and \textit{T2 relaxation} respectively.

>>>>

\subsection{Basic Principles}
\subsubsection{Spin and Magnetic Moment}
Spin is an intrinsic property of elementary particles, and can be either positive or negative. The spin quantum number  $s$ is defined as $ s = \frac{n}{2}$, where $n \in \{0,1,2,...\}$. Therefore $s$ is a quantized value in multiples of $\frac{1}{2}$. Protons, electrons and neutrons possess spin. Nuclei with an odd number of protons or neutrons have non-zero nuclear spins; while those with an even number of protons and neutrons have no spin. Particles with opposite spins can pair up to produce an overall spin of zero. Hydrogen nuclei, which makes up 63$\%$ the human body, possess spins of $\frac{1}{2}$. The spin of a particle has a magnetic moment vector that behaves like a magnet with north and south poles (Figure \ref{fig:spin}).

@@

There are more protons with spins with parallel orientation than anti-parallel, due to the increased energy stability. This number of difference is called spin excess and the magnetic moments of the excess spins add to form the net magnetization vector $M_z$. When the spin system is at equilibrium, net magnetization lies along the z-axis, parallel to $B_0$.

@@

\subsubsection{Relaxation}
Transverse magnetization is introduced into the system that oscillates at radio frequency (RF) to modify the net magnetization, and this is called the RF pulse. When exposing the nuclear spin system to the energy of the RF pulse, the equilibrium of the spin system is disturbed. The RF pulse is designed to introduce energy E such that the net spin direction orients at $90^\circ$. The net magnetization now is denoted as $M_{xy}$ as it aligns with the xy-plane instead of the z-axis. Individual spins now precess in phase with the RF pulse, and it gives rises to the MR signal in the receiver coil.

With the termination of the RF pulse, the nuclei in the excited state will return to the equilibrium condition. This process is called \textit{relaxation}. 
The energy transmitted decays rapidly by two relaxation processes:  spin-lattice and spin-spin relaxations. These are responsible for \textit{T1 relaxation} and \textit{T2 relaxation} respectively.
